---
title: "Introduction to MRCCC"
author:
  - name: Bitan Sarkar
  - name: Yang Ni
package: MRCCC
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Introduction to MRCCC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Introduction

Cell-cell communication is usually inferred from the co-expression of a
ligand in one cell type and its receptor in another. Co-expression, however,
is not evidence of a causal signal: shared regulators, tissue composition and
technical factors all induce correlations between cell types without any
signalling taking place. `r BiocStyle::Biocpkg("MRCCC")` implements MR-CCC, a
Bayesian Mendelian randomization framework that uses cis-eQTL genotypes as
instrumental variables for ligand expression in a sender cell type and for
receptor expression in a receiver cell type, and asks whether the
genetically predicted ligand level, modulated by the genetically predicted
receptor level, shifts the activity of a downstream pathway in the receiver.

The unit of analysis is a ligand-receptor-pathway triplet. For each triplet
the package returns a posterior inclusion probability (PIP) that summarises
the evidence for a causal signal, together with effect estimates, credible
intervals, posterior sign probabilities, the receptor level at which the
ligand effect reverses sign, and the Monte Carlo and convergence diagnostics
needed to trust these numbers. Across many triplets, the PIPs can be
thresholded with Bayesian false discovery rate control.

```{r load}
library(MRCCC)
```

# The model

For donor $i$, let $X_i$ be ligand expression in the sender cell type, $Z_i$
receptor expression in the receiver cell type, and $Y_i$ pathway activity in
the receiver cell type. Let $G_i$ and $H_i$ be the cis-eQTL genotypes that
instrument the ligand and the receptor, and $V_i$ a vector of shared donor
covariates. The first stage predicts each exposure from its instruments,
$X_i^* = E[X_i \mid G_i, V_i]$ and $Z_i^* = E[Z_i \mid H_i, V_i]$. The second
stage models the outcome as
$$
Y_i = \mu + \gamma\,(\beta_X X_i^* + \beta_{XZ} X_i^* Z_i^*) + \beta_Z Z_i^*
    + \alpha_Y^\top V_i + \varepsilon_{Yi},
$$
where $\gamma \in \{0, 1\}$ is a spike-and-slab inclusion indicator on the
causal block $(\beta_X, \beta_{XZ})$. Because $X^*$ and $Z^*$ depend on
genotype only, they are unaffected by the unmeasured confounders that
distort a naive regression of $Y$ on $X$ and $Z$. All three stages are
estimated jointly by a blocked Gibbs sampler.

The posterior inclusion probability $\mathrm{PIP} = P(\gamma = 1 \mid
\text{data})$ is the communication score. The receptor-modulated ligand
effect is $\beta_X + \beta_{XZ} z$, which changes sign at $\tau = -\beta_X /
\beta_{XZ}$: when $\tau$ lies inside the observed receptor range, the ligand
activates the pathway in some donors and suppresses it in others. The model
requires $X$, $Z$ and $Y$ to be centred; `mr_ccc()` does this internally and
reports effects on a standardised scale (per standard deviation of $X$ and
$Z$, in standard deviations of $Y$) alongside the raw scale.

# Simulating a triplet

`simulate_mrccc()` generates one triplet from the working model with an
unmeasured confounder that loads on the ligand, the receptor and the
pathway. The default generates 300 donors, five instruments per gene, three
covariates, and a positive interaction.

```{r simulate}
sim <- simulate_mrccc(n = 300, beta_X = 0.3, beta_XZ = 0.3, beta_Z = 0.5,
                      conf_strength = 0.7, seed = 1)
str(sim[c("X", "Z", "Y", "G", "H", "V")])
sim$truth
```

# Fitting the model

`mr_ccc()` validates the inputs, centres $X$, $Z$ and $Y$, runs the Gibbs
sampler, and pools the retained draws. The chain length below is kept short
so that this document builds quickly; the section on chain length discusses
the settings appropriate for reported results.

```{r fit}
fit <- mr_ccc(sim$X, sim$Z, sim$Y, sim$G, sim$H, sim$V,
              n_iter = 4000, burn_in = 500, seed = 1)
fit
```

`summary()` prints the PIP with its Monte Carlo standard error, the
standardised effects with credible intervals, and the convergence
diagnostics, and returns the assembled table invisibly.

```{r summary}
tab <- summary(fit)
tab
```

# Input from a SummarizedExperiment

Donor-level (pseudo-bulk) expression is commonly stored in a
`SummarizedExperiment`, with genes as rows and donors (samples) as columns.
`se_row()` extracts one gene from such an object as a named numeric vector
that can be passed to `mr_ccc()` as `X`, `Z` or `Y`. A
`SummarizedExperiment` that contains exactly one gene can also be passed to
`mr_ccc()` directly, in which case its first assay row is used. The chunk
below stores the simulated ligand expression in a one-gene
`SummarizedExperiment` and refits the model from it, with a short chain to
keep this document fast.

```{r se-input}
m <- rbind(LIG1 = as.numeric(sim$X))
colnames(m) <- paste0("donor", seq_len(nrow(sim$G)))
se <- SummarizedExperiment::SummarizedExperiment(assays = list(logcounts = m))
x_se <- se_row(se, "LIG1")
fit_se <- mr_ccc(x_se, sim$Z, sim$Y, sim$G, sim$H, sim$V,
                 n_iter = 1500, burn_in = 300, seed = 1)
fit_se$pip
```

# Posterior summaries

## Credible intervals

```{r ci}
credible_intervals(fit)
credible_intervals(fit, level = 0.9, standardized = FALSE)
```

## Sign probabilities of the interaction

The prior on $(\beta_X, \beta_{XZ})$ is a spike-and-slab mixture, so the
marginal posterior of $\beta_{XZ}$ mixes draws taken while $\gamma = 0$
(concentrated near zero) with draws taken while $\gamma = 1$. The marginal
sign probabilities combine the evidence that there is an effect with the
evidence about its direction; the conditional probabilities describe the
direction given that the ligand acts on the pathway. Both are informative.

```{r signs}
sign_probabilities(fit)
```

## Sign-reversal threshold

`sign_reversal()` computes $\tau = -\beta_X / \beta_{XZ}$ from paired draws
and reports it in standard-deviation units of receptor expression relative
to the mean. The element `p_in_range` is the posterior probability that
$\tau$ lies inside the observed receptor range, which determines whether the
reversal is realised in the donor population.

```{r tau}
sr <- sign_reversal(fit, Z_observed = sim$Z)
sr[c("tau_median", "tau_lower", "tau_upper", "Z_range_std", "p_in_range")]
```

In this simulation the true threshold is $-0.3 / 0.3 = -1$ raw unit of $Z$,
which corresponds to `r round(-1 / sd(sim$Z), 2)` standard deviations.

## Effect curve

`plot_effect_curve()` draws the posterior mean of $\beta_X + \beta_{XZ} z$
over the observed receptor range with a pointwise credible band built from
paired draws.

```{r curve, fig.cap = "Receptor-modulated ligand effect with a 95 percent pointwise credible band."}
plot_effect_curve(fit, Z_observed = sim$Z)
```

# Choosing the chain length

The PIP is the mean of an autocorrelated binary chain, so its Monte Carlo
standard error (MCSE) is estimated by non-overlapping batch means within each
chain and combined across chains in quadrature. The effective sample size
(ESS) is the number of independent draws that would give the same precision.

```{r mcse}
c(pip = fit$pip, mcse = fit$pip_mcse, ess = fit$pip_ess)
```

When the PIP lies within two MCSEs of the discovery threshold (0.5 by
default), `mr_ccc()` issues a warning: the discovery status of the triplet
is not resolved at the current chain length, and either `n_iter` or
`n_chains` should be increased. The warning is a property of the Monte Carlo
approximation, not of the data; a longer run narrows the MCSE until the
status is resolved or the PIP is shown to sit genuinely at the threshold.

The short chain used in this document is adequate for illustration only. For
reported results, substantially longer chains should be used; the analysis in
the paper pools four chains of 100,000 iterations.

Chains alternate their starting value of $\gamma$ and initialise
$(\beta_X, \beta_{XZ})$ from a normal distribution with standard deviation
`init_scale`, so they begin at dispersed points. The Gelman-Rubin statistic
$\hat R$ is then reported for $\beta_X$, $\beta_{XZ}$, $\gamma$ and the joint
log-likelihood in `fit$diagnostics`; it is `NA` if a single chain was run,
because $\hat R$ compares the variance between chains with the variance
within them and there is nothing to compare. This is why the default is
`n_chains = 4`.

Pooling chains is not more expensive than lengthening one. Per-chain Monte
Carlo standard errors combine in quadrature, so $C$ chains of `n_iter`
iterations carry the same standard error as one chain of $C \times$ `n_iter`;
the difference is that several chains also yield $\hat R$.

```{r chains}
fit2 <- mr_ccc(sim$X, sim$Z, sim$Y, sim$G, sim$H, sim$V,
               n_iter = 2000, burn_in = 500, n_chains = 2, seed = 2)
fit2$diagnostics
```

# Screening many triplets

In a full analysis `mr_ccc()` is applied to every candidate triplet and the
PIPs are collected in a vector. For a declared set $S$ the Bayesian false
discovery rate is
$$
\mathrm{bFDR}(S) = |S|^{-1} \sum_{j \in S} (1 - \mathrm{PIP}_j),
$$
the posterior expected proportion of false discoveries. `bayes_fdr()` returns
the bFDR of each nested top-$k$ set ranked by PIP, and
`bfdr_threshold_set()` selects the largest set whose bFDR does not exceed a
target level.

```{r bfdr}
pips <- c(0.98, 0.95, 0.91, 0.84, 0.72, 0.61, 0.48, 0.30, 0.12, 0.05)
data.frame(pip = pips, bfdr = bayes_fdr(pips),
           selected_10 = bfdr_threshold_set(pips, alpha = 0.10),
           selected_20 = bfdr_threshold_set(pips, alpha = 0.20))
```

# Session information

```{r session}
sessionInfo()
```
